Planet Evolution

### 2014-03-11 ms.tex
Testing...

\documentclass{emulateapj}
% To do 
% 1) consistency between p and P for pressure
% 2) Signs in fluxes... and overall consistency (is Fint everywhere it
%    needs to be?)

@@

\author{
%
Greg Novak\altaffilmark{1} \&
%
David S. Spiegel\altaffilmark{2}
}

@@

\begin{figure*}
  \centering
  % from regime_plot() in ./gsn.py
  \includegraphics[width=0.9\textwidth]{f3.pdf}\\
  \caption{Phase diagram of the relevant regimes.  {\it Top:} The
    sound speeds given just use the non-relativistic idea gas formula.
    There are no relativistic corrections (so it can be greater than
    $c$), I don't worry about the how the mean molecular weight
    changes with temperature, or about how the sound speed changes
    when, e.g. the electrons are degenerate. {\it Bottom:} The dashed
    lines are just reference points and do not divide physical
    regions.}
  \label{fig:eos-regimes}
\end{figure*}

@@

\begin{figure*}
  \centering
  % from plot_single_channel() in structure/python/rc.py
  \includegraphics[width=0.45\textwidth]{taurc-vs-text-high.pdf}
  \hfill
  \includegraphics[width=0.45\textwidth]{taurc-vs-text-low.pdf}
  \caption{Left, optical depth of radiative/convective boundary for a
    single channel of stellar irradiation, deposited high in the
    atmosphere ($\tau\simeq 0.1$).  The x axis, characterizes the
    temperature profile of the atmosphere, with shallow profiles on
    near zero and steep profiles near 1.  On the right, the optical
    depth of the radiative-convective boundary for the case where the
    stellar irradiation is deposited low in the atmosphere.  The jumps
    in the value of $\tau_{rc}$ indication places where a deep
    radiative layer is formed.}
  \label{fig:detached-radiative-zones}
\end{figure*}

\begin{figure*}
  \centering
  % from plot_1chan_2d() in structure/python/rc.py
  \includegraphics[width=0.49\textwidth]{../plots/10/09/fbon-flux-contour-low.pdf}
  \hfill
  \includegraphics[width=0.49\textwidth]{../plots/10/09/fbon-flux-contour-high.pdf}
  \caption{Contours of radiative/convective boundary vs. structure of
    atmosphere (along the x axis) and ratio of external flux to
    internal flux (y axis).  On the left, depositing energy low in
    the atmosphere, on the right, depositing energy high.  The
    behavior is relatively smooth when depositing energy high in the
    atmosphere.  A discontinuity develops when depositing energy low
    in the atmosphere, and a fixed amount of energy has a larger
    effect.  }
  \label{fig:atmosphere-vs-flux}
\end{figure*}

@@

\begin{figure}
  \centering
  % from plot_1chan_2d() in structure/python/rc.py
  \includegraphics[width=1.0\columnwidth]{../plots/10/09/fbon-k-contour.pdf}
  \caption{If we hold the ratio of external flux to internal flux
    constant and ask what happens to the rad/conv boundary as the
    structure of the atmosphere (x axis) and depth of energy deposit
    (y axis, high is high, low is low), then we can see the ``fold''
    develop as the energy is deposited deeper and deeper in the
    atmosphere.}
  \label{fig:atmosphere-vs-depth}
\end{figure}

@@

\begin{figure*}
  \centering
% from plot_1chan_2d() in structure/python/rc.py
\includegraphics[width=0.32\textwidth]{../plots/10/09/flux-k-contour-fbon-0p2.pdf}
\hfill
\includegraphics[width=0.32\textwidth]{../plots/10/09/flux-k-contour-fbon-0p7.pdf}
\hfill
\includegraphics[width=0.32\textwidth]{../plots/10/09/flux-k-contour-fbon-0p9.pdf}
\caption{Optical depth of the rad/conv transition for three different
  atmosphere structures, from shallow temp profile (left) to steep
  temp profile (right).  The x axis gives the ratio of external flux
  to internal flux (small ratio on the left, large ratio on the right)
  while the y axis gives the depth at which the energy is deposited
  (low for low, high for high).  One sees that depositing the energy
  deeper has a larger effect, and a ``fold'' structure develops for
  atmospheres with shallow temperature profiles.}
\label{fig:flux-vs-depth}
\end{figure*}

@@

\begin{figure*}
% from plot_grav_1chan() in structure/python/rc.py
\includegraphics[width=0.45\textwidth]{../plots/10/09/grav-1chan-fbon-0p5-k-0p1.pdf}
\includegraphics[width=0.45\textwidth]{../plots/10/09/grav-1chan-fbon-0p5-k-10.pdf}
\caption{Contours of surface gravity as a function of effective
  temperature of external radiation (x axis) and effective temperature
of of planet's cooling radiation (y axis).  On the left, for energy
deposited deep in the atmosphere, on the right, high in the
atmosphere.   One can see that at fixed surface gravity (along a
contour), increasing Text causes Tint to fall in a well-defined
fashion, with somewhat interesting behavior around $Text \simeq Tint$
when depositing energy deep in the atmosphere.}
\label{fig:surf-grav-vs-text-and-tint}
\end{figure*}

@@

\begin{figure}
% from plot_grav_1chan() in structure/python/rc.py
\includegraphics[width=0.45\textwidth]{../plots/10/09/grav-1chan-fbon-0p2-k-0p1.pdf}
\caption{Contours of surface gravity as a function of effective
  temperature of external radiation (x axis) and effective temperature
  of of planet's cooling radiation (y axis), when depositing energy
  deep in the atmosphere.  The same as the previous figure, except
  that the structure of the atmosphere is different, the temperature
  profile being shallower.  One sees that the contours of surface
  gravity develop an ``overbite'' as Text approaches Tint.  This would
  indicate that for fixed surface gravity, Tint is no longer a
  single-valued function of Text.  I'm not sure what this means.  Does
  stability determine which solution is selected?}
\label{fig:surf-grav-vs-text-and-tint-shallow-temp-profile}
\end{figure}

@@

@@

We thank Adam Burrows and Ivan Hubeny for useful conversations.  GSN
was supported by the ERC European Research Council under the Advanced
Grant Program Num 267399-Momentum, and by the National Science
Foundation under Grant No. PHY11-25915 via the Kavli Institute for
Theoretical Physics.  DSS gratefully acknowledges a fellowship from
the AMIAS group.

>>>>

\documentclass{emulateapj}
% DSS test 10 March 2014

% To do 
% 1) consistency between p and P for pressure
% 2) Signs in fluxes... and overall consistency (is Fint everywhere it
%    needs to be?)

@@

\author{
%
Greg Novak\altaffilmark{1},
%
David S. Spiegel\altaffilmark{2}
}

@@

\begin{figure*}
  \centering
  % from regime_plot() in ./gsn.py
  \includegraphics[width=0.9\textwidth]{f3}\\
  \caption{Phase diagram of the relevant regimes.  {\it Top:} The
    sound speeds given just use the non-relativistic idea gas formula.
    There are no relativistic corrections (so it can be greater than
    $c$), I don't worry about the how the mean molecular weight
    changes with temperature, or about how the sound speed changes
    when, e.g. the electrons are degenerate. {\it Bottom:} The dashed
    lines are just reference points and do not divide physical
    regions.}
  \label{fig:eos-regimes}
\end{figure*}

@@

\begin{figure*}
  \centering
  % from plot_single_channel() in structure/python/rc.py
  \includegraphics[width=0.45\textwidth]{taurc-vs-text-high}
  \hfill
  \includegraphics[width=0.45\textwidth]{taurc-vs-text-low}
  \caption{Left, optical depth of radiative/convective boundary for a
    single channel of stellar irradiation, deposited high in the
    atmosphere ($\tau\simeq 0.1$).  The x axis, characterizes the
    temperature profile of the atmosphere, with shallow profiles on
    near zero and steep profiles near 1.  On the right, the optical
    depth of the radiative-convective boundary for the case where the
    stellar irradiation is deposited low in the atmosphere.  The jumps
    in the value of $\tau_{rc}$ indication places where a deep
    radiative layer is formed.}
  \label{fig:detached-radiative-zones}
\end{figure*}

\begin{figure*}
  \centering
  % from plot_1chan_2d() in structure/python/rc.py
  \includegraphics[width=0.49\textwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/fbon-flux-contour-low}
  \hfill
  \includegraphics[width=0.49\textwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/fbon-flux-contour-high}
  \caption{Contours of radiative/convective boundary vs. structure of
    atmosphere (along the x axis) and ratio of external flux to
    internal flux (y axis).  On the left, depositing energy low in
    the atmosphere, on the right, depositing energy high.  The
    behavior is relatively smooth when depositing energy high in the
    atmosphere.  A discontinuity develops when depositing energy low
    in the atmosphere, and a fixed amount of energy has a larger
    effect.  }
  \label{fig:atmosphere-vs-flux}
\end{figure*}

@@

\begin{figure}
  \centering
  % from plot_1chan_2d() in structure/python/rc.py
  \includegraphics[width=1.0\columnwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/fbon-k-contour}
  \caption{If we hold the ratio of external flux to internal flux
    constant and ask what happens to the rad/conv boundary as the
    structure of the atmosphere (x axis) and depth of energy deposit
    (y axis, high is high, low is low), then we can see the ``fold''
    develop as the energy is deposited deeper and deeper in the
    atmosphere.}
  \label{fig:atmosphere-vs-depth}
\end{figure}

@@

\begin{figure*}
  \centering
% from plot_1chan_2d() in structure/python/rc.py
\includegraphics[width=0.32\textwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/flux-k-contour-fbon-0p2}
\hfill
\includegraphics[width=0.32\textwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/flux-k-contour-fbon-0p7}
\hfill
\includegraphics[width=0.32\textwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/flux-k-contour-fbon-0p9}
\caption{Optical depth of the rad/conv transition for three different
  atmosphere structures, from shallow temp profile (left) to steep
  temp profile (right).  The x axis gives the ratio of external flux
  to internal flux (small ratio on the left, large ratio on the right)
  while the y axis gives the depth at which the energy is deposited
  (low for low, high for high).  One sees that depositing the energy
  deeper has a larger effect, and a ``fold'' structure develops for
  atmospheres with shallow temperature profiles.}
\label{fig:flux-vs-depth}
\end{figure*}

@@

\begin{figure*}
% from plot_grav_1chan() in structure/python/rc.py
\includegraphics[width=0.45\textwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/grav-1chan-fbon-0p5-k-0p1}
\includegraphics[width=0.45\textwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/grav-1chan-fbon-0p5-k-10}
\caption{Contours of surface gravity as a function of effective
  temperature of external radiation (x axis) and effective temperature
of of planet's cooling radiation (y axis).  On the left, for energy
deposited deep in the atmosphere, on the right, high in the
atmosphere.   One can see that at fixed surface gravity (along a
contour), increasing Text causes Tint to fall in a well-defined
fashion, with somewhat interesting behavior around $Text \simeq Tint$
when depositing energy deep in the atmosphere.}
\label{fig:surf-grav-vs-text-and-tint}
\end{figure*}

@@

\begin{figure}
% from plot_grav_1chan() in structure/python/rc.py
\includegraphics[width=0.45\textwidth]{\string~/Dropbox/GregDave/paper1_planet_evolution/plots/10/09/grav-1chan-fbon-0p2-k-0p1}
\caption{Contours of surface gravity as a function of effective
  temperature of external radiation (x axis) and effective temperature
  of of planet's cooling radiation (y axis), when depositing energy
  deep in the atmosphere.  The same as the previous figure, except
  that the structure of the atmosphere is different, the temperature
  profile being shallower.  One sees that the contours of surface
  gravity develop an ``overbite'' as Text approaches Tint.  This would
  indicate that for fixed surface gravity, Tint is no longer a
  single-valued function of Text.  I'm not sure what this means.  Does
  stability determine which solution is selected?}
\label{fig:surf-grav-vs-text-and-tint-shallow-temp-profile}
\end{figure}

@@

\section{Maximum Entropy of Solutions}

I find myself thinking again about the maximum entropy of solutions of
finite radius.  If the entire solution is on an adiabat, what can I
say about that solution?  Consider first the usual case where
electrons are fully degenerate and provide pressure while protons are
non-degenerate and provide entropy.

EOS:
\begin{equation}
  P = \frac{\hbar}{m_e} n^{5/3}
\end{equation}
HSE:
\begin{equation}
  P_c = \frac{G M^2}{R^4}
\end{equation}
together:
\begin{equation}
  R = \frac{\hbar^2}{m_e m_p^{5/3} G M^{1/3}}
\end{equation}
which gives
\begin{equation}
  P_c = \frac{G^5 M^{10/3} m_e^4 m_p^{20/3}}{\hbar^8}
\end{equation}
I can also get the escape velocity of the planet
\begin{equation}
  v_{esc} = \frac{G M^{2/3} m_e^{1/2} m_p^{5/6}}{\hbar}
\end{equation}
Now I know that $T\propto n^{2/3}$ for an adiabat.  More specifically
\begin{equation}
  \psi = \log \frac{n}{n_{q,e}} = \log \left(\frac{T}{T_F}\right)^{-3/2}
\end{equation}
\begin{equation}
  n = n_{q,e} e^\psi
\end{equation}
then
\begin{equation}
  n = \left(\frac{m_e k T}{\hbar^2}\right)^{3/2} e^\psi
\end{equation}
and
\begin{equation}
  kT = \frac{\hbar^2}{m_e} e^{-2\psi/3} n^{2/3}
\end{equation}
I want to know something at the surface.  Do I want to know the
thermal velocity or the Fermi velocity?  Do I want to know the
velocity of the electrons or the protons?  It's not clear to me.
\begin{equation}
  v_{Fe} = \sqrt{\frac{k T_F}{m_e}} = \frac{\hbar n^{1/3}}{m_e}
\end{equation}
\begin{equation}
  v_{te} = \sqrt{\frac{k T}{m_e}}
\end{equation}
\begin{equation}
  v_{tp} = \sqrt{\frac{k T}{m_p}}
\end{equation}

Now I would like estimates of the density and temperature at the
surface.  One definition of the surface is $P=0$ but that gives $\rho
= T = 0$.  I want the point where the $T$ profile goes off of an
adiabat.  This will happen at least above the $\tau=1$ surface,
perhaps deeper.  Let's say that it happens at $N$ optical depths.
\begin{equation}
  \delta R = \frac{N}{\rho \kappa}
\end{equation}
and HSE gives
\begin{equation}
  \frac{P_s - 0}{\delta R} = \frac{G M \rho}{R^2}
\end{equation}
so 
\begin{equation}
  P_s = \frac{G M N}{\kappa R^2} 
  = \frac{N g}{\kappa} 
  = \frac{G^3M^{5/3} N m_e^2 m_p^{10/3}}{\kappa \hbar^4}
\end{equation}
which gives the surface number density
\begin{equation}
  n_s = \frac{G^{9/5} M N^{3/5} m_e^{9/5} m_p^2}{\kappa^{3/5} \hbar^{18/5}}
\end{equation}
and the surface temperature
\begin{equation}
  k T_s = \frac{e^{-2\psi/3} G^{6/5} M^{2/3} N^{2/3} m_e^{1/5} m_p^{4/3}}{\kappa^{2/5}\hbar^{2/5}}
\end{equation}

These allow me to write the velocities at the surface.  Repeating the
escape velocity and putting in the surface temperature/density for the others:
\begin{equation}
  v_{esc} = \frac{G M^{2/3} m_e^{1/2} m_p^{5/6}}{\hbar}
\end{equation}
(note that this looks like a ratio involving the Chandra mass)
\begin{equation}
  v_{Fe} = \frac{G^{3/5} M^{1/3} N^{1/5} m_p^{2/3}}
  {\kappa^{1/5} \hbar^{1/5} m_e^{2/5}}
\end{equation}

\begin{equation}
  v_{te} = \frac{e^{-\psi/3}G^{3/5} M^{1/3} N^{1/5} m_p^{2/3}}
  {\kappa^{1/5} \hbar^{1/5} m_e^{2/5}}
\end{equation}

\begin{equation}
  v_{tp} = \frac{e^{-\psi/3}G^{3/5} M^{1/3} N^{1/5} m_p^{1/6} m_e^{1/10}}
  {\kappa^{1/5} \hbar^{1/5}}
\end{equation}

So $v_{te}$ is the same as $v_{Fe}$ suppressed by the degeneracy
prameter, and $v_{tp}$ is the same as $v_{te}$, suppressed by the
square root of the electron proton mass ratio.  But the escape
velocity scales differently.  Form the ratio
\begin{equation}
  \frac{v_{Fe}}{v_{esc}} = \frac{N^{1/5} \hbar^{4/5}}{\kappa^{1/5}
    G^{2/5} M^{1/3} m_e^{1/2} m_p^{1/6}}
\end{equation}
This is $6 \times 10^{-11}$ for a Jupiter mass, or unity for 0.4
grams.  That is to say that all relevant velocities are much less than
the escape velocity.

\subsection{Lane-Emden Equation}
Taking the solutions of the Lane-Emden equation and putting the
dimensional parts back in, we have:

\begin{equation}
  % K+W 19.3
  P = K \rho ^\gamma = K \rho^{1+\frac{1}{n}}
\end{equation}
\begin{equation}
  % K+W 19.9
  R = \frac{z_n}{A}  
\end{equation}
\begin{equation}
  A = \sqrt{\frac{4\pi G}{(n+1) K}}\rho_c^{\frac{n-1}{2n}}
\end{equation}
\begin{equation}
  %K+W 19.20
  \rho_c = \frac{M z_n^3}{4\pi R^3 D_n}
\end{equation}
\begin{equation}
  % Take value from Table 19.1
  D_n = \left( -z^2 \frac{dw}{dz}\right)_{z=z_n}
\end{equation}
where $w$ and $z$ are the dimensionless variables in which the LE
equation is written, $z_n$ is the value of $z$ where $w=0$ (the
surface of the object)

Then we may finally write 
\begin{equation}
  R = z_n 
  M^{\frac{1-n}{3-n}}
  (n+1)^{\frac{n}{3-n}}
  K^{\frac{n}{3-n}}
  (4 \pi D_n)^{\frac{n-1}{3-n}}
  (4 \pi G)^{\frac{n}{n-3}}
\end{equation}

Now what is $K$?  Dispense with all unitless coefficient of order
unity, but keep everything that appears in an exponent.  It's always
the case that protons provide the entropy (thus fix the temperature)
and electrons provide the pressure.  I want to write the EOS in terms
of density and entropy.  Consider three cases in turn.  First, protons
and electrons are both degenerate (so that the entropy is given above,
and is not the Sakur-Tetrode equation).  Then
\begin{equation}
  P = \frac{\hbar^2}{m_e m_p^{5/3}} \left( 1+\frac{kT}{kT_F} \right) \rho^{5/3} 
\end{equation}
with 
\begin{equation}
  k T_F = \frac{\hbar^2 n^{2/3}}{m_e}
\end{equation}
and 
\begin{equation}
  kT = \frac{\hbar^2 n^{2/3} \sigma}{m_p}
\end{equation}
so 
\begin{equation}
  P = \frac{\hbar^2}{m_e m_p^{5/3}}
\left(1 + \frac{m_e \sigma}{m_p} \right)
\rho^{5/3}
\end{equation}
which is valid for $\sigma < 1$

Second case: electrons degenerate, protons non-degenerate.  Then the
Sakur-Tetrode equation gives
\begin{equation}
  kT = \frac{\hbar^2}{m_p} n^{2/3} 
\exp\left(\frac{2\sigma}{3} - \frac{5}{3}\right)
\end{equation}
and the EOS is
\begin{equation}
  P = \frac{\hbar^2}{m_e m_p^{5/3}}
\left[1 + \frac{m_e}{m_p} 
\exp\left(\frac{2\sigma}{3} - \frac{5}{3}\right) \right]
\rho^{5/3}
\end{equation}
which is valid for $1 < \sigma < 11$

Third case: electrons and protons non-degenerate.  The relation
between entropy and temperature is the same, so the EOS gives.
\begin{equation}
  P = \frac{\hbar^2}{m_p^{8/3}} 
\exp\left(\frac{2\sigma}{3} - \frac{5}{3}\right) 
\rho^{5/3}
\end{equation}
which is valid for $11 < \sigma$.

@@
